A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-11 tex/codatadp.tex
more

\section{Theory of (Co)datatypes}
\label{sec:theory-of-co-datatypes}

@@

Finally, some codatatypes are so degenerate that they have infinite values
even though they are finite. The simplest example is
\[\keyw{datatype}\;\, \ty{a} \,=\, \const{A}(\ty{a})\]
This is a finite codatatype, of cardinality one, whose single value is
infinite: $\mu\vthinspace a.\; A(a)$. Other specimens are
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-\jot]
             \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{unit},\: \ty{c},\: \ty{b},\: \ty{c})
\end{aligned}
\]
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$. We call such types \emph{corecursive singletons}.
For the decision procedure, it will be important to detect those finite
codatatypes.

@@

\begin{lemma}
If the domain interpreting a corecursive codatatype is finite, it must be $1$.
\end{lemma}

\begin{proof}
TODO.
\qed
\end{proof}

An immediate consequence by the distinctness axiom is that these types have
necessarly one constructor and that all arguments have types of cardinality 1.
%
Sometimes, a type might be a corecursive singleton only in some models. For
example, if \ty{unit} is an uninterpreted type, \ty{b} and \ty{c} will be
singleton types precisely in the models that interpret \ty{unit} as a
singleton.

@@

We introduced a decision procedure for datatypes and codatatypes and implemented
it in the SMT solver CVC4. Our main theoretical contribution has been the
addition of codatatypes. These are not very difficult
per se, but there are tricky corner cases to take care of if we care about
completeness (which we do). On the practical side, we obtained interesting
results on benchmarks generated from Isabelle theories, including some rather
%mind-boggling
puzzling bisimulation proofs.

This work is part of a larger program that aims at enriching automatic
provers with more high-level features and reducing the gap between automatic and
interactive theorem proving. As future work, we want to try out CVC4 and its
(co)datatypes for higher-order model finding in Isabelle, as an
alternative to the counterexample generator Nitpick
\cite{blanchette-nipkow-2010}. We also see some opportunities to enrich CVC4
with recursive and corecursive functions. Finally, it might be possible to go
further in terms of supporting nested and mixed (co)recursion and quantified
formulas over (co)datatypes in SMT solvers.

\def\ackname{Acknowledgment}
\paragraph{\ackname.}
We owe a great debt to Clark Barrett and Cesare Tinelli for datatype case blah
blah blah.
%
Our present and former bosses, Viktor Kuncak, Stephan Merz, Tobias Nipkow,
Cesare Tinelli, and Christoph Weidenbach, have either encouraged the research on
codatatype or at least tolerated it, both of which we are thankful for.
%
Andrei Popescu helped clarify our thoughts about axiomatizing codatatypes.
Andreas Lochbihler and Peter Gammie shared their private codatatype-based
theories with us so we could include them in the benchmarks.
%
Blanchette's research was partially supported by the Deutsche
Forschungs\-gemein\-schaft (DFG) project
\relax{Hardening the Hammer} (grant Ni\,491\slash 14-1).

>>>>

\section{The Theory of (Co)datatypes}
\label{sec:the-theory-of-co-datatypes}

@@

Finally, some codatatypes are so degenerate that they have infinite values
even though they are finite. The simplest example is
\[\keyw{datatype}\;\, \ty{a} \,=\, \const{A}(\ty{a})\]
This is a finite codatatype, of cardinality one, whose single value is
infinite: $\mu\vthinspace a.\; A(a)$. Other specimens are
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-\jot]
             \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{unit},\: \ty{c},\: \ty{b},\: \ty{c})
\end{aligned}
\]
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$. We call such types \emph{corecursive singletons}.
For the decision procedure, it will be important to detect those finite
codatatypes. Sometimes, a type might be a corecursive singleton only in some
models. For example, if \ty{unit} is an uninterpreted type, \ty{b} and \ty{c}
will be singleton types precisely in the models that interpret \ty{unit} as a
singleton.

@@

\begin{lemma}
If the domain interpreting a corecursive codatatype is finite, it must be a
singleton. Furthermore, such a type has necessarly a single constructor and
all its arguments have types that are interpreted by singletons.
\end{lemma}

\begin{proof}
  * if two (either directly or indirectly) corecursive constructors
    $\const{C}$ and $\const{D}$, it is
    possible to encode infinitely many alternation patterns,
    e.g. $\const{C}(\const{D}(\const{C}(\const{C}(\ldots))))$;
    hence no finite models

  * if one corecursive and one noncorecursive, then possibility to create
    objects of arbitrary depths, e.g. $\const{C}(\ldots(\const{C}(\const{D}))\ldots)$,
    and hence no finite models

  * leaves us with the case of a single corecursive constructor

  * if any of the noncorecursive arguments has cardinality greater than one,
    then it is possible to encode alternation patterns using it,
    e.g. $\const{C}(0,\: \const{C}(1,\: \const{C}(0,\: \const{C}(0,\: \ldots))))$.

  * otherwise, by the coinduction principle, there is only one value that can
    exist

%  * assume for simplicity no indirect recursion, but this does not radically
%    change the argument
%  * assume there are at least two values built with C.
%    they must be different at some point, e.g.
%    C(0, C(1, C(0, ...))) and C(0, C(1, C(1, ...)))
%    can use that to create infinitely alternation patterns
%  * leaves us with the case of a single constructor
% * since we are looking at a specific model, we can assume all ordinary
%   types are finite datatypes with nullary constructors corresponding to their
%   elements
% * if two
% * either there is only one ``path'' from ctr to itself
%   or at least two;
\qed
\end{proof}

@@

We introduced a decision procedure for the ground theory of datatypes and
codatatypes and implemented it in the SMT solver CVC4. Our main theoretical
contribution has been the support for codatatypes. These are not very
difficult per se, but there are tricky corner cases to take care of if we care
about completeness (which we do).
%% alliteration
On the practical side, we obtained
interesting results on benchmarks generated from Isabelle theories, including
some rather
%mind-boggling
puzzling bisimulation proofs.

This work is part of a larger program that aims at enriching automatic
provers with more high-level features and reducing the gap between automatic and
interactive theorem proving. As future work, we want to try out CVC4 and its
(co)datatypes for higher-order model finding in Isabelle, as an
alternative to the counterexample generator Nitpick
\cite{blanchette-nipkow-2010}. We also see some opportunities to enrich
SMT solvers with recursive and corecursive functions. Finally, it might be
possible to go further in terms of supporting nested and mixed (co)recursion
and quantified formulas over (co)datatypes in solvers.

\def\ackname{Acknowledgment}
\paragraph{\ackname.}
We owe a great debt to Clark Barrett and Cesare Tinelli for datatype case blah
blah blah. \textbf{TODO FIXME}
%
Our present and former bosses, Viktor Kuncak, Stephan Merz, Tobias Nipkow,
Cesare Tinelli, and Christoph Weidenbach, have either encouraged the research on
codatatype or at least endured
%tolerated?
it, both of which we are thankful for.
%
Andreas Lochbihler and Peter Gammie shared their private
theories with us so we could include them in the benchmarks.
Andrei Popescu helped clarify our thoughts regarding the axiomatization of
codatatypes. Dmitriy Traytel took part in discussions about special
cases.
%
Blanchette's research was partially supported by the Deutsche
Forschungs\-gemein\-schaft (DFG) project
\relax{Hardening the Hammer} (grant Ni\,491\slash 14-1).
